Illustrating Model Properties

### 2024-04-22 manuscript.qmd
mv history to models.qmd, add totals to data table

```{python}
#| echo: false
import re

@@

def highlight_bool(
  column: pd.Series,
  c_false: str = COLORS["green"],
  c_true: str = COLORS["red"],
) -> list[str]:
  """Make cell read (`False`) or green (`True`)."""
  n = len(column)
  trues = pd.Series([c_true] * n)
  falses = pd.Series([c_false] * n)
  colors = trues.where(column, falses)
  return colors.map(lambda x: f"color: {x}")

@@

In @fig-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.

@@

```{python}
#| echo: false
#| label: fig-burnin-history 
#| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
from collections import namedtuple

def custom(x, pos):
  return "{:.1f}".format(x/1000)

nrows, ncols = 1, 3
label_map = {
  "acor_times": "autocorrelation [steps]",
  "accept_fracs": "acceptance fraction [%]",
  "max_log_probs": r"max. log-likelihood [$\times 10^3$]",
}

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(
  nrows=nrows,
  ncols=ncols,
  aspect_ratio=1.,
  width=full,
))

fig, axes = plt.subplots(
  nrows=nrows,
  ncols=ncols,
  sharex=True,
)

for model in ["contra", "bilateral", "midline"]:
  history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
  for i, column in enumerate(history.columns):
    history.plot(y=column, ax=axes[i], label=model)
    axes[i].autoscale(enable=True, tight=True, axis="x")
    axes[i].set_ylabel(label_map[column])
    axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
    axes[i].set_xlabel(r"steps [$\times 10^3$]")

axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[2].yaxis.set_major_formatter(ticker.FuncFormatter(custom))
plt.show()
```

@Fig-burnin-history is only a temporary placeholder for now.

@@

      if model == "contra":
        scenario.side = "contra"
        scenario.is_uni = True

      print(scenario)

@@

```{python}
#| echo: false
#| label: tbl-data-strat
#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
data = raw.copy()
data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
early_midext_data = (data.loc[data[COL.t_stage] == "early", COL.midext] == True).mean()
late_midext_data = (data.loc[data[COL.t_stage] == "late", COL.midext] == True).mean()
grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])

lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
total = grouped.count()[lnl_cols]["max_llh", "contra"]
idx = total.index.rename(["Mid. ext.", "T-cat.", "ipsi III"])

num_involved = grouped.sum()[lnl_cols]["max_llh", "contra"]
percent_involved = num_involved / total
total.index = idx
num_involved.index = idx
percent_involved.index = idx

involved = num_involved.join(
  100 * percent_involved,
  rsuffix=" (%)",
).sort_index(axis="columns")
involved.columns = pd.MultiIndex.from_product(
  [["I", "II", "III", "IV"],
   ["n", "%"]],
  names=["LNL", ""]
)
(
  involved
  .reset_index()
  .style
  .format(precision=2)
  .apply(right_align, subset=involved.columns[0:])
  .apply(highlight_bool, subset=["Mid. ext.", "ipsi III"])
  .apply(highlight_t_stage, subset=["T-cat."])
  .hide()
)
```

>>>>

```{python}
#| echo: false

import re

@@

def highlight(column: pd.Series, mapping: dict) -> list[str]:
  """Color based on `mapping`."""
  colors = column.map(mapping)
  return colors.map(lambda x: f"color: {x}")

def highlight_bool(
  column: pd.Series,
  c_false: str = COLORS["green"],
  c_true: str = COLORS["red"],
) -> list[str]:
  """Make cell read (`False`) or green (`True`)."""
  return highlight(column, mapping={True: c_true, False: c_false})

@@

In @fig-model-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.

@@

{{< embed model.qmd#fig-model-burnin-history >}}

@fig-model-burnin-history is only a temporary placeholder for now.

@@

      if model == "contra":
        scenario.side = "contra"
        scenario.is_uni = True
      else:
        scenario.side = "ipsi"
        scenario.is_uni = False

@@

```{python}
#| echo: false
#| label: tbl-data-strat
#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
lnl_cols = get_lnl_cols("contra", lnls=["I", "II", "III", "IV"])
num_ipsi_inv = raw[get_lnl_cols("ipsi")].sum(axis="columns")

contra_inv = raw[lnl_cols].copy()
contra_inv.columns = contra_inv.columns.droplevel([0,1])
contra_inv["t_stage"] = raw[COL.t_stage].apply(lambda x: "early" if x <= 2 else "advanced")
contra_inv["ipsi"] = num_ipsi_inv.map(lambda x: str(x) if x <= 1 else "≥ 2")
contra_inv["midext"] = raw[COL.midext]

grouped = contra_inv.groupby(by=["t_stage", "ipsi", "midext"])
num_involved = grouped.sum()
total = grouped.count()
percent_involved = num_involved / total
idx = total.index.rename(["T-cat.", "ipsi", "Mid. ext."])

total.index = idx
num_involved.index = idx
percent_involved.index = idx

involved = num_involved.join(
  100 * percent_involved,
  rsuffix=" (%)",
).sort_index(axis="columns")
involved.columns = pd.MultiIndex.from_product(
  [["I", "II", "III", "IV"],
   ["n", "%"]],
  names=["LNL", ""]
)
involved["total", "n"] = total["I"]
(
  involved
  .reset_index()
  .sort_values("ipsi", ascending=True)
  .sort_values("T-cat.", ascending=False)
  .style
  .format(precision=2)
  .apply(right_align, subset=involved.columns[0:])
  .apply(highlight_t_stage, subset=["T-cat."])
  .apply(highlight_bool, subset=["Mid. ext."])
  .apply(highlight, subset=["ipsi"], mapping={
    "0": COLORS["green"],
    "1": COLORS["orange"],
    "≥ 2": COLORS["red"],
  })
  .hide()
)
```

### 2024-04-22 model.qmd
mv history to models.qmd, add totals to data table

>>>>

# Sampling History

For bookeeping, we also show how different quantities evolve during the burnin of the sampling. For example, the ratio of proposed samples that were accepted or the autocorrelation time.

```{python}
#| echo: false
#| label: fig-model-burnin-history 
#| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
from collections import namedtuple
import pandas as pd

def custom(x, pos):
  return "{:.1f}".format(x/1000)

nrows, ncols = 1, 3
label_map = {
  "acor_times": "autocorrelation [steps]",
  "accept_fracs": "acceptance fraction [%]",
  "max_log_probs": r"max. log-likelihood [$\times 10^3$]",
}

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(
  nrows=nrows,
  ncols=ncols,
  aspect_ratio=1.,
  width=17,
))

fig, axes = plt.subplots(
  nrows=nrows,
  ncols=ncols,
  sharex=True,
)

for model in ["contra", "bilateral", "midline"]:
  history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
  for i, column in enumerate(history.columns):
    history.plot(y=column, ax=axes[i], label=model)
    axes[i].autoscale(enable=True, tight=True, axis="x")
    axes[i].set_ylabel(label_map[column])
    axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
    axes[i].set_xlabel(r"steps [$\times 10^3$]")

axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[2].yaxis.set_major_formatter(ticker.FuncFormatter(custom))
plt.show()
```
